Hierarchical Time Series Forecasting in Emergency Medical Services

### 2023-07-18 manuscript/main.tex
ensemble 2 removed from table 3

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{todonotes,mathtools,bm,amsmath}
\mathtoolsset{showonlyrefs}
\usepackage{mathpazo}
\setlength{\parindent}{0cm}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Annals of Emergency Medicine}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Hierarchical Time Series Forecasting in Emergency Medical Services},
  pdfauthor={Bahman Rostami-Tabar; Rob J. Hyndman},
  pdfkeywords={healthcare, emergency services, forecast
reconciliation, ambulance demand, regression},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Accurate forecasts of ambulance demand are crucial inputs when planning
and deploying staff and fleet. Such demand forecasts are required at
national, regional and sub-regional levels, and must take account of the
nature of incidents and their priorities. These forecasts are often
generated independently by different teams within the organization. As a
result, forecasts at different levels may be inconsistent, resulting in
conflicting decisions and a lack of coherent coordination in the
service. To address this issue, we exploit the hierarchical and grouped
structure of the demand time series, and apply forecast reconciliation
methods to generate both point and probabilistic forecasts that are
coherent and use all the available data at all levels of disaggregation.
The methods are applied to daily incident data from an ambulance service
in Great Britain, from October 2015 to July 2019, disaggregated by
nature of incident, priority, managing health board, and control area.
We use an ensemble of forecasting models, and show that the resulting
forecasts are better than any individual forecasting model. We validate
the forecasting approach using time-series cross-validation.
\end{abstract}

\begin{keyword}
    healthcare \sep emergency services \sep forecast
reconciliation \sep ambulance demand \sep 
    regression
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi

@@

A failure to match available resources to demand in Emergency Medical
Services (EMS) results in patient flow problems, with serious
consequences for patients, staff and the entire care system
\citep{ekstrom2015forecasting, ROSTAMITABAR20221197}. Demand forecasting
in EMS helps service planners to avoid the mismatch, potentially
providing massive savings in costs and lives, and leading to better
patient outcomes. Accurate daily demand forecasting enables planners and
decision makers to manage resources to meet anticipated patients,
reconfigure units, and redeploy staff and vehicles as necessary.

Demand forecasts at EMS are typically required at multiple levels of an
organization to inform various planning and decision-making processes
\citep{hulshof2012taxonomic}. There are some planning process at the
national level (strategic and long-term) such as workforce resource
planning and budgeting; sub-national, regional or healthcare level
(tactical and medium-term) such as temporary capacity expansions,
resource sharing and staff-shift scheduling; and hospital or station
level (operational and short-term) such as planning rosters for staff
and ambulance deployment. Demand forecasts might also be required at
different levels for a specific area of interest such as the nature of
demand or the priority level. Moreover, the time series data in EMS has
an inherent hierarchical and grouped structure to support such
forecasting requirements. Demand for emergency medical services at the
country level can be disaggregated in a geographical hierarchy into
sub-national, regions, health boards, and stations/hospitals, or divided
into groups such as the nature of incidents or demand priority.
Therefore, using forecasting methodologies that account for hierarchical
and/or grouped structures of time series in EMS is a natural fit.

However, despite a large number of studies dedicated to forecasting for
EMS
\citep{mingliterature2022, gul2020exhaustive, ibrahim2016modeling, wargon2009systematic},
the hierarchical data structure has been largely ignored, and the main
focus has been on producing independent forecasts at a single level.
Generating independent forecasts can result in a lack of consistency and
coordination, and therefore leads to less effective planning and
decision making. With hierarchical forecasting, plans at any level are
based on coherent forecasts and therefore can be aligned. Implementing
and sustaining improvements in EMS require alignments and coordination
between different stakeholders, without which teams operate in isolation
leading to conflicts, duplication work, rework, or work that runs
counter to the overall goal to improve the quality of delivery service.
Hierarchical forecasting framework can be used as a tool to improve
coordination between teams across the care services at the national,
sub-national, regional and local levels. The hierarchical forecasting
approaches not only create consistent forecasts, but are usually also
more accurate than the independent (base) forecasts
\citep{hyndman2011optimal}. To our knowledge, there has been no previous
research involving hierarchical and grouped forecasting in the entire
field of forecasting for healthcare management.

In this paper, we address this gap by investigating the application of
hierarchical forecasting approaches in the EMS using daily time series
of verified incidents from 2015 to 2020 in a major ambulance service in
Great Britain. The data has hierarchical and grouped structures, with
hierarchies at the national, control (i.e.~sub-national), health board
(i.e.~regional) levels, as well as groups by priority and nature of
incidents. We produce consistent point forecasts and forecast
distributions for all levels, which is critical for an effective
planning and associated risk management. We compare the point and
probabilistic forecast accuracy of the independent forecasts, bottom-up
and optimal reconciliation approaches. We first generate
independent/base forecasts using Exponential Smoothing State Space
(ETS), Generalized Linear Model (GLM), Poisson regression, a simple
empirical distribution and an ensemble method, followed by applying
bottom-up and optimal reconciliation approaches. Forecast performance is
assessed by the Mean Squared Scaled Error (MSSE) for point forecasts and
Continuous Ranked Probability Scores (CRPS) for the probabilistic
forecasts. This paper complies with reproducibility principles
\citep{stodden2013best, boylan2015reproducibility}, and can be applied
to any healthcare service (e.g., emergency department, primary or social
care) subject to the time series having a hierarchical and/or grouped
structure.

@@

Emergency medical services (EMS) are a critical component in the
delivery of urgent medical care to communities. An effective service
delivery requires accurate resource planning that generally relies on
demand forecasts at operational, tactical and strategic levels.

There is a substantial number of studies on the application of time
series forecasting in the Emergency Medical Services. For example,
\citet{ibrahim2016modeling} provide an extensive review of the models
used in forecasting call volume arrivals. Another important area is
related to forecasting ambulance demand. Although the definition of
demand might not be always clearly stated, this is typically referring
to a situation where a physical resource has been deployed to respond to
an incident. This might be also called \emph{attended incidents}.
Another demand related variable is verified incidents; these are all
incidents that require an action: either by sending a physical vehicle,
responding via the Clinical Support Desk, requesting an external
provider to respond to it, or forwarding it to other channels such as
police, firefighters or general practitioners. Our study is aligned with
this stream of the literature. Another similar area that has received
considerable attention is Emergency Department forecasting; we refer
interested readers to \citet{mingliterature2022},
\citet{gul2020exhaustive} and \citet{wargon2009systematic} for extensive
reviews of the relevant literature. Although crucial to EMS performance,
\citet{aringhieri2017emergency} state that demand forecasting has
received limited research attention in the EMS context. In this section,
we provide a brief review of studies on forecasting ambulance demand in
EMS.

@@

\citet{sasaki2010using} develop a multivariable regression model to
estimate future EMS demands. In addition to the historical demand, the
population census for different age groups and counts of the number of
companies employing more than five people are included in the
regression. The census variables describe groups who are more likely to
need an ambulance. A stepwise ordinary least squares regression analysis
is used for estimating the parameter and generating forecast. The only
performance measure reported in this study is \(R^2\), which is not an
effective measure of forecast accuracy \citep[p457]{armstrong01}. The
research design of this study is not rigorous and the study is not
reproducible. \citet{vile2012predicting} explore using a Singular
Spectrum Analysis (SSA) method to generate forecasts of the EMS demand
at the national level for 7-day, 14-day, 21-day and 28-day forecast
horizons using data provided by an ambulance service in Great Britain.
The performance of this approach is compared to Auto-Regressive
Integrated Moving Average (ARIMA) and Holt-Winters time series methods
using Root Mean Squared Error (RMSE). They concluded that point
forecasts generated by SSA are more accurate for longer-term, but that
ARIMA and Holt-Winters performance is superior for shorter-term
horizons. \citet{vile2016time} further develop a decision support system
to integrate forecasts generated by SSA. However, the study does not
compare and contrast the performance of forecasting methods based on
utility measures such as cost, resource utilization or response time.
The tool contains options that allow generating forecasts at various
levels of granularity, however, it ignores the hierarchical and grouped
relationships structure, preventing aligned decision making and
coordination.

\citet{9659837} investigate forecasting EMS demand in a high
spatio-temporal resolution of 1km\(^2\) spatial regions and 1-hr time
intervals using total incidents in Oslo, Norway, from 1 January 2015 to
11 February 2019. They used multi-layer perceptron (MLP) and long
short-term memory (LSTM) models to forecast the EMS demand, and compare
the results to simple aggregation methods and baselines. The point
forecast accuracy is evaluated using Mean Absolute Error (MAE) and Mean
Squared Error (MSE), and the forecast distribution is measured by
Categorical Cross-Entropy. They found that Neural Network models
performed better in producing point forecasts, while a distribution
baseline method based on spatial distribution of the incidents across
all time steps provided more accurate forecast distributions.
\citet{zhou2016predictinglit} proposed three methods based on Gaussian
mixture models, kernel density estimation, and kernel warping to predict
hourly data 4 weeks ahead for a 1km\(^2\) spatial region. Two years of
incidents from Toronto, Canada (years 2007 and 2008 with 391,296 events)
and Melbourne, Australia (years 2011 and 2012 with 696,975 events) were
used to build the model and examine the performance on test data using
mean negative log likelihood. They show that forecasts generated by the
proposed methods were significantly more accurate than the current
industry practice (a simple averaging formula).
\citet{grekousis2019will} investigated the combination of spatial
analysis methods with data mining techniques based on an improved
Hungarian algorithm and a MLP neural network to identify the most likely
locations of future emergency events. The proposed approach was tested
using data of 2851 events attended by the EMS in Athens, Greece, over 24
weeks. They showed that 23\% of real emergency events lie within 50
meters of the predicted ones and nearly 70\% of the real emergency
events lie no further than 150 meters away, which is rather accurate
given the granularity of the problem at the city level.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Current studies ignore the inherent hierarchical and/or grouped
  structure of the time series data, and the relationship between series
  at different levels of hierarchy. This may result in incoherent
  forecasts leading to misaligned planning and decision making. While
  the hierarchical forecasting methodology has been developed and
  applied in various domains over the past 10 years
  \citep{panagiotelis2022probabilistic}, it has never been explored in
  this area.
\item
  Current research is mainly concerned with generating point forecasts
  at a single level of hierarchy. There is a lack of studies considering
  the entire forecast distribution of daily ambulance demand for the
  whole hierarchy to inform the decision-making process and to better
  represent the uncertainty of future demand, providing a risk
  management tool for planners.
\item
  Reproducibility is still a major challenge in EMS forecasting, as it
  is unlikely that any reader can reproduce prior studies without the
  help of the authors of those papers.
\item
  Another limitation is related to the generated forecasts not being on
  the sample space of non-negative counts. Since actual ambulance counts
  cannot be negative or non-integer, ambulance demand forecast
  distributions should reflect the data. Of course, point forecasts
  represent means, so they should be non-negative, but may be
  non-integer. While this might not be an issue when producing forecasts
  at a single level, producing non-negative count forecasts in a
  hierarchical/grouped structure is challenging and requires further
  investigation in the future.
\end{enumerate}

This paper concerns the problem of hierarchical forecasting in EMS and
generates and evaluates both point and probabilistic forecast across
different levels of the hierarchy, hence addressing some important gaps
identified in the literature.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
\mathtoolsset{showonlyrefs}
\setlength{\parindent}{0cm}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal of Service Research}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Hierarchical Time Series Forecasting in Emergency Medical Services},
  pdfauthor={Bahman Rostami-Tabar; Rob J. Hyndman},
  pdfkeywords={healthcare, emergency services, forecast
reconciliation, ambulance demand, regression},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Accurate forecasts of ambulance demand are crucial inputs when planning
and deploying staff and fleet. Such demand forecasts are required at
national, regional, and sub-regional levels, and must take account of
the nature of incidents and their priorities. These forecasts are often
generated independently by different teams within the organization. As a
result, forecasts at different levels may be inconsistent, resulting in
conflicting decisions and a lack of coherent coordination in the
service. To address this issue, we exploit the hierarchical and grouped
structure of the demand time series and apply forecast reconciliation
methods to generate both point and probabilistic forecasts that are
coherent and use all the available data at all levels of disaggregation.
The methods are applied to daily incident data from an ambulance service
in Great Britain, from October 2015 to July 2019, disaggregated by
nature of incident, priority, managing health board, and control area.
We use an ensemble of forecasting models and show that the resulting
forecasts are better than any individual forecasting model. We validate
the forecasting approach using time series cross validation.
\end{abstract}

\begin{keyword}
    healthcare \sep emergency services \sep forecast
reconciliation \sep ambulance demand \sep 
    regression
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

A failure to match available resources to demand in Emergency Medical
Services (EMS) results in patient flow problems, with serious
consequences for patients, staff, and the entire care system
\citep{ekstrom2015forecasting, ROSTAMITABAR20221197}. Demand forecasting
in EMS helps service planners to avoid the mismatch, potentially
providing massive savings in costs and lives, and leading to better
patient outcomes. Accurate daily demand forecasting enables planners and
decision-makers to manage resources to meet anticipated patients,
reconfigure units, and redeploy staff and vehicles as necessary.

Demand forecasts at EMS are typically required at multiple levels of an
organization to inform various planning and decision-making processes
\citep{hulshof2012taxonomic}. There are some planning processes at the
national level (strategic and long-term) such as workforce resource
planning and budgeting; sub-national, regional, or healthcare level
(tactical and medium-term) such as temporary capacity expansions,
resource sharing; and hospital or station level (operational and
short-term) such as planning rosters for staff and ambulance deployment.
Demand forecasts might also be required at different levels for a
specific area of interest such as the nature of demand or the priority
level. Moreover, the time series data in EMS has an inherent
hierarchical and grouped structure to support such forecasting
requirements. Demand for emergency medical services at the national
level can be disaggregated in a geographical hierarchy into
sub-national, regions, health boards, and stations/hospitals, or divided
into groups such as the nature of incidents or demand priority.
\textcolor{teal}{Forecasts produced at both higher and lower levels of hierarchies are necessary for effective decision-making in EMS. For example, control area EMS forecasts can inform strategic decisions about how to allocate limited resources to lower levels, such as health boards and stations/hospitals. At the lower levels, hospitals or ambulance stations could use such forecasts to plan for staffing and resource allocation, ambulance dispatching, staff-to-shift assignment, staff rescheduling based on the anticipated volume and priority and nature of incidents. Additionally, generating forecasts at lower levels could potentially improve the accuracy of the high-level forecasts, by providing more detailed information on the nature and priority of incidents. This could help to identify patterns in demand that may not be apparent at the higher level. Therefore, employing forecasting techniques that consider the hierarchical and/or grouped patterns of time series in EMS aligns naturally, offering the possibility to enhance forecast accuracy and facilitate coordination.}

However, despite a large number of studies dedicated to forecasting for
EMS
\citep{mingliterature2022, gul2020exhaustive, ibrahim2016modeling, wargon2009systematic},
the hierarchical data structure has been largely ignored, and the main
focus has been on producing independent (base) forecasts at a single
level. Generating independent forecasts can result in a lack of
consistency and coordination, and therefore leads to less effective
planning and decision making.
\textcolor{teal}{For example, suppose the annual budget for the whole organization is to be allocated to each area health board based on demand within the corresponding area. If the forecasts are incoherent, there is a mismatch between the total resources to be allocated, and the resources allocated to each area. Further, suppose the budget also needs to take account of the nature of incidents that occur within each area, with more money allocated for some types of incidents than others. Then we need forecasts of the demand disaggregated by nature of incidents and health board, but these data are often too noisy to forecast directly. Forecast reconciliation addresses these problems by ensuring the forecasts are coherent across all aggregated and disaggregated series (avoiding mismatches), and by using the signal in the aggregated data to allow forecasting of highly disaggregated data (allowing better targetting of the budget).}

With hierarchical forecasting, plans at any level are based on coherent
forecasts and therefore can be aligned. Implementing and sustaining
improvements in EMS require alignments and coordination between
different stakeholders, without which teams operate in isolation leading
to conflicts, duplication work, rework, or work that runs counter to the
overall goal to improve the quality of delivery service. Hierarchical
forecasting framework can be used as a tool to improve coordination
between teams across the care services at the national, sub-national,
regional and local levels. The hierarchical forecasting approaches not
only create consistent forecasts but are usually also more accurate than
the independent (base) forecasts \citep{hyndman2011optimal}. To our
knowledge, there has been no previous research involving hierarchical
and grouped forecasting in the entire field of forecasting for
healthcare management.

In this paper, we address this gap by investigating the application of
hierarchical forecasting approaches in the EMS using daily time series
of attended incidents from 2015 to 2019 in a major ambulance service in
Great Britain. The data has hierarchical and grouped structures, with
hierarchies at the national, control (i.e.~sub-national), and health
board (i.e.~regional) levels, as well as groups by priority and nature
of incidents. We produce consistent point forecasts and forecast
distributions for all levels, which is critical for an effective
planning and associated risk management. We compare the point and
probabilistic forecast accuracy of the independent forecasts, bottom-up
and optimal reconciliation approaches. We first generate
independent/base forecasts using Exponential Smoothing State Space
(ETS), Poisson regression using Generalized Linear Model (GLM) and
tscount (TSGLM), a simple empirical distribution and an ensemble method,
followed by applying bottom-up and optimal reconciliation approaches.
Forecast performance is assessed by the Mean Absolute Scaled Error
(MASE) and Mean Squared Scaled Error (MSSE) for point forecasts and
Continuous Ranked Probability Scores (CRPS) for the probabilistic
forecasts. This paper complies with reproducibility principles
\citep{stodden2013best, boylan2015reproducibility}. We provide the R
codes for the proposed models and benchmarks. Therefore, they can be
applied to any healthcare service (e.g., emergency department, primary
or social care) subject to the time series having a hierarchical and/or
grouped structure.

@@

Emergency medical services (EMS) are a critical component in the
delivery of urgent medical care to communities. An effective service
delivery requires accurate resource planning that generally relies on
demand forecasts at operational, tactical, and strategic levels. There
is a substantial number of studies on the application of time series
forecasting in the Emergency Medical Services. For example,
\citet{ibrahim2016modeling} provide an extensive review of the models
used in forecasting call volume arrivals. Another important area is
related to forecasting ambulance demand. Although the definition of
demand might not be always clearly stated, this is typically referring
to a situation where a physical resource has been deployed to respond to
an incident. This might be also called \emph{attended incidents}.
Another demand related variable is verified incidents; these are all
incidents that require an action: either by sending a physical vehicle,
responding via the Clinical Support Desk, requesting an external
provider to respond to it, or forwarding it to other channels such as
police, firefighters, or general practitioners. Our study is aligned
with this stream of literature. Another similar area that has received
considerable attention is Emergency Department forecasting; we refer
interested readers to \citet{mingliterature2022},
\citet{gul2020exhaustive}, and \citet{wargon2009systematic} for
extensive reviews of the relevant literature. In this section, we
provide a brief review of studies on forecasting ambulance demand in
EMS.

@@

\citet{sasaki2010using} develop a multivariable regression model to
estimate future EMS demands. In addition to the historical demand, the
population census for different age groups and counts of the number of
companies employing more than five people are included in the
regression. The census variables describe groups who are more likely to
need an ambulance. A stepwise ordinary least squares regression analysis
is used for estimating the parameter and generating forecasts. The only
performance measure reported in this study is \(R^2\), which is not an
effective measure of forecast accuracy \citep[p457]{armstrong01}. The
research design of this study is not rigorous and the study is not
reproducible. \citet{vile2012predicting} explore using a Singular
Spectrum Analysis (SSA) method to generate forecasts of the EMS demand
at the national level for 7-day, 14-day, 21-day, and 28-day forecast
horizons using data provided by an ambulance service in Great Britain.
The performance of this approach is compared to Auto-Regressive
Integrated Moving Average (ARIMA) and Holt-Winters time series methods
using Root Mean Squared Error (RMSE). They concluded that point
forecasts generated by SSA are more accurate for longer-term, but that
ARIMA and Holt-Winters performance is superior for shorter-term
horizons. \citet{vile2016time} further develop a decision support system
to integrate forecasts generated by SSA. However, the study does not
compare and contrast the performance of forecasting methods based on
utility measures such as cost, resource utilization and response time.
The tool contains options that allow generating forecasts at various
levels of granularity; however, it ignores the structure of the
hierarchical and grouped relationships, preventing aligned decision
making and coordination. \citet{al2021empirical}
\textcolor{teal}{utilises real data from the Welsh Ambulance Service to explore the forecast accuracy of four forecasting approaches: ARIMA, Holt Winters, Multiple Regression, and Singular Spectrum Analysis (SSA) in predicting call volume demand accurately. The aim is to compare these approaches with the current method across various planning horizons (7 days, 30 days, and 90 days) for both total call volume and category-specific demand. Forecast accuracy performance is evaluated using root mean square error (RMSE) and mean absolute percentage error (MAPE) are calculated. The findings indicate that ARIMA performs the best in predicting weekly and monthly demand. However, when it comes to long-term demand, the SSA method proves to be the most effective.}
\citet{ibrahim2016modeling}
\textcolor{teal}{conducted a case study to assess the effectiveness of multiple forecasting methods: the multiplicative univariate forecasting model (MU), univariate mixed-effects model (ME), bivariate mixed-effects model (BME1), and bivariate mixed-effects model (BME2). Real-life call centre data was utilised to forecast for periods of 1, 7, and 14 days ahead, using only a limited dataset of 42 days. The performance of these forecasting methods was evaluated using two metrics: RMSE for point forecasts and coverage probability for the 95\% prediction interval. The findings indicate that the ME consistently produces the most accurate point forecasts. On the other hand, BME1 and BME2 demonstrate superior coverage probabilities when forecasting for one day or one week ahead. For a two-week leading period, MU shows better coverage probability.}

\citet{9659837} investigate forecasting EMS demand in a high
Spatio-temporal resolution of 1km\(^2\) spatial regions and 1-hr time
intervals using total incidents in Oslo, Norway, from 1 January 2015 to
11 February 2019. They used multi-layer perceptron (MLP) and long
short-term memory (LSTM) models to forecast the EMS demand, and compare
the results to simple aggregation methods and baselines. The point
forecast accuracy is evaluated using Mean Absolute Error (MAE) and Mean
Squared Error (MSE), and the forecast distribution is measured by
Categorical Cross-Entropy. They found that Neural Network models
performed better in producing point forecasts, while a distribution
baseline method based on the spatial distribution of the incidents
across all time steps provided more accurate forecast distributions.
\citet{zhou2016predictinglit} proposed three methods based on Gaussian
mixture models, kernel density estimation, and kernel warping to predict
hourly data 4 weeks ahead for a 1km\(^2\) spatial region. Two years of
incidents from Toronto, Canada (years 2007 and 2008 with 391,296 events)
and Melbourne, Australia were used to build the model and examine the
performance on test data using mean negative log-likelihood. They show
that forecasts generated by the proposed methods were significantly more
accurate than the current industry practice (a simple averaging
formula). \citet{grekousis2019will} investigated the combination of
spatial analysis methods with data mining techniques based on an
improved Hungarian algorithm and a MLP neural network to identify the
most likely locations of future emergency events. The proposed approach
was tested using data from 2851 events attended by the EMS in Athens,
Greece, over 24 weeks. They showed that 23\% of real emergency events
lie within 50 meters of the predicted ones and nearly 70\% of the real
emergency events lie no further than 150 meters away, which is rather
accurate given the granularity of the problem at the city level.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Current studies ignore the inherent hierarchical and/or grouped
  structure of the time series data, and the relationship between series
  at different levels of hierarchy. While the hierarchical forecasting
  methodology has been developed and applied in various domains over the
  past 10 years \citep{panagiotelis2022probabilistic}, it has never been
  explored in this area.
\item
  Current research is mainly concerned with generating point forecasts
  at a single level of hierarchy. There is a lack of studies considering
  the entire forecast distribution of daily ambulance demand for the
  whole hierarchy to better represent the uncertainty of future demand,
  providing a risk management tool for planners.
\item
  Reproducibility is still a major challenge in EMS forecasting, as it
  is unlikely that any reader can reproduce prior studies without the
  help of the authors of those papers.
\item
  Another limitation is related to the generated forecasts not being on
  the sample space of non-negative counts. Since actual ambulance counts
  cannot be negative or non-integer, ambulance demand forecast
  distributions should reflect the data. Of course, point forecasts
  represent means, so they should be non-negative, but may be
  non-integer. While this might not be an issue when producing forecasts
  at a single level, producing non-negative count forecasts in a
  hierarchical/grouped structure is challenging and requires further
  investigation in the future.
\end{enumerate}

\begin{landscape}\begin{table}[!h]

\caption{Summary of some studies in forecasting FOR AMBULANCE SERVICES}
\centering
\resizebox{\linewidth}{!}{
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{lr>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{8em}ll}
\toprule
Reference & Year & Variable & Horizon & Method & Metric & Probabilistic & Reconciliation\\
\midrule
Current study & 2023 & Ambulance demand & 84 days & Naïve;  Exponential Smoothing State Space (ETS), Poisson regression using Generalized Linear Model (GLM) and tscount (TSGLM), a simple empirical distribution and an ensemble method & MASE, MSSE, CRPS & YES & YES\\
Al-Azzani et al. & 2021 & Call volume & 7 , 30 ,  90 days & ARIMA, Holt Winters, Multiple Regression, and Singular Spectrum Analysis & RMSE, MAPE & NO & NO\\
Haugsbø et al. & 2021 & Ambulance demand in Spatio-temporal & 1hour & MLP, LSTM & MAE, MSE, Cross-Entropy & Yes & NO\\
Grekousis et al. & 2019 & Locations of incidents & 1 hour & MLP and Hungarian algorithm & RMSE & NO & NO\\
Ibrahim et al. & 2016 & Call volume & 1, 7 , 14 days & multiplicative univariate forecasting, univariate mixed-effects, bivariate mixed-effects model, and bivariate mixed-effects & RMSE, prediction interval coverage & Partial & NO\\
Vile et al. & 2012 & Ambulance demand & 7, 14, 21,  28 day & Singular Spectrum Analysis, ARIMA,  Holt-Winters & RMSE & No & NO\\
Sasaki et al. & 2010 & Ambulance demand & 5 years & OLS regression & R\textasciicircum{}2 & NO & NO\\
\bottomrule
\end{tabular}}
\end{table}
\end{landscape}
